---
title: "Flow matching as an alternative to EDM diffusion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Flow matching as an alternative to EDM diffusion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      purl = FALSE)
```

`RDiffPuter` ships with two interchangeable generative backbones:

- `method = "edm"` (default): the EDM diffusion model from
  [Karras et al. (2022)](https://arxiv.org/abs/2206.00364), trained with
  weighted denoising MSE and sampled with a 2nd-order Heun ODE solver.
  This is the formulation in the original DiffPuter paper.
- `method = "flow"`: a flow-matching velocity field from
  [Lipman et al. (2023)](https://arxiv.org/abs/2210.02747), trained with
  plain regression on a linear interpolant between Gaussian noise and data,
  sampled with explicit Euler. Typically needs ~5–10 sampling steps for
  quality comparable to 50 EDM steps.

This vignette shows how to switch between them and what to expect.

```{r}
library(RDiffPuter)
library(torch)
device <- if (cuda_is_available()) "cuda" else
          if (torch::backends_mps_is_available()) "mps" else "cpu"
```

## A small mixed-types data set

```{r}
set.seed(2)
n <- 300
truth <- data.frame(
  age   = pmax(18, rnorm(n, 40, 12)),
  inc   = exp(rnorm(n, 10, 0.6)),
  edu   = factor(sample(c("hs", "ba", "ma", "phd"), n, replace = TRUE)),
  reg   = factor(sample(c("N", "S", "E", "W", "C"), n, replace = TRUE))
)
data <- truth
for (col in names(data)) {
  miss <- sample.int(n, n * 0.25)
  data[miss, col] <- NA
}
```

## EDM diffusion

The original recipe. `num_steps = 50`, `num_resamplings = 20`,
`num_trials = 20` are the recommended defaults; we shrink them here just so
the vignette completes quickly.

```{r}
edm_run <- diffputer_em(
  data, method = "edm",
  max_iter   = 2, epochs = 200, hidden_dim = 128,
  batch_size = 64, patience = 50,
  num_steps  = 30, num_resamplings = 10, num_trials = 8,
  device     = device, validation_data = truth, verbose = TRUE
)
```

## Flow matching

Same call, just `method = "flow"`. Defaults branch automatically to
`num_steps = 10`, `num_resamplings = 5`, `num_trials = 10` — substantially
fewer net evaluations per imputation.

```{r}
flow_run <- diffputer_em(
  data, method = "flow",
  max_iter   = 2, epochs = 200, hidden_dim = 128,
  batch_size = 64, patience = 50,
  device     = device, validation_data = truth, verbose = TRUE
)
```

## Single-step (no inner re-flow)

For the quickest possible flow-imputation, set
`resample_strategy = "single"` to drop the inner re-flow loop. With
`num_trials > 1` you still benefit from Monte Carlo averaging across
trials.

```{r}
flow_fast <- diffputer_em(
  data, method = "flow", resample_strategy = "single",
  max_iter   = 2, epochs = 200, hidden_dim = 128,
  batch_size = 64, patience = 50,
  num_trials = 16,
  device     = device, validation_data = truth, verbose = TRUE
)
```

## Comparing wall clock and quality

```{r}
compare <- rbind(
  edm       = edm_run$history,
  flow      = flow_run$history,
  flow_fast = flow_fast$history
)
compare
```

## What flow matching trades off

The conditional-imputation construction in DiffPuter (Theorem 1 of the
ICLR 2025 paper) is derived for the diffusion case. The flow-matching
analogue — mixing the linear-interpolant forward path on observed entries
with the model's velocity step on missing entries — is a sound *heuristic*
adaptation, but it does not have the same exactness guarantee. Empirically
the gap is small for tabular imputation; if your application requires
provable correctness with respect to the paper's analysis, stick with
`method = "edm"`. If you need fast inference and mostly care about
practical quality, `method = "flow"` is the better choice.
