---
title: "Getting started with RDiffPuter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RDiffPuter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      purl = FALSE)
```

`RDiffPuter` ports the diffusion-based imputer
[DiffPuter](https://github.com/hengruizhang98/DiffPuter) (Zhang et al., ICLR 2025)
to R. It accepts an arbitrary `data.frame` whose missing entries are encoded
as `NA` and returns a completed `data.frame` with the same schema. Continuous
and categorical columns are handled automatically and training runs on CPU,
CUDA, or Apple Silicon (MPS).

## Installation

```{r}
# install.packages("torch")  # one-time backend install
# torch::install_torch()
# remotes::install_github("mneunhoe/RDiffPuter")
library(RDiffPuter)
library(torch)
```

## A numeric example

We draw 200 rows from a multivariate normal, blank out 30% of the entries at
random, and run two iterations of the EM loop.

```{r}
set.seed(1)
n <- 200
truth <- data.frame(a = rnorm(n), b = rnorm(n, 1, 0.7), c = rnorm(n, -1, 1.3))
data <- truth
mask <- matrix(runif(n * 3) < 0.3, n, 3)
data[mask] <- NA

# Pick the best device available.
device <- if (cuda_is_available()) "cuda" else if (
  torch::backends_mps_is_available()) "mps" else "cpu"

result <- diffputer_em(
  data,
  max_iter        = 2,
  epochs          = 200,
  patience        = 50,
  hidden_dim      = 128,
  batch_size      = 64,
  num_steps       = 20,
  num_trials      = 8,
  num_resamplings = 5,
  device          = device,
  verbose         = TRUE
)

head(result$imputed)
```

`result$imputed` is a `data.frame` with the same columns as `data`. Observed
entries are preserved exactly; only the cells that were `NA` are filled in.

## Inspecting the run

```{r}
print(result)
plot(result$history$loss, type = "b",
     xlab = "EM iteration", ylab = "M-step training loss")
```

## Power-user path: train once, impute many times

The high-level call above is convenient, but if you want to train the
diffusion model once and reuse it on multiple data sets that share a schema,
work directly with [`data_transformer`](`?data_transformer`) and
[`diffputer_trainer()`](`?diffputer_trainer`):

```{r}
transformer <- data_transformer$new()
transformer$fit(data)          # learns means, sds, level vocabularies
encoded <- transformer$transform(data)

trained <- diffputer_trainer(
  encoded$x, hidden_dim = 128, epochs = 200,
  batch_size = 64, patience = 50, device = device
)

# Impute the original data:
filled <- impute(trained, data, transformer,
                 num_trials = 8, num_steps = 20, num_resamplings = 5)

# Impute a fresh batch (out-of-sample) without retraining:
new_data <- truth[1:50, ]
new_data$a[c(3, 9, 14)] <- NA
filled_new <- impute(trained, new_data, transformer,
                     num_trials = 8, num_steps = 20, num_resamplings = 5)
```

## Saving and loading

```{r}
save_diffputer(result, "models/run1")
restored <- load_diffputer("models/run1")
filled <- impute(restored$model, data, restored$transformer)
```
