---
title: "Imputing mixed numeric and categorical columns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Imputing mixed numeric and categorical columns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      purl = FALSE)
```

This vignette walks through an end-to-end imputation of a small mixed
numeric/categorical data frame, including a comparison of the two
categorical encodings RDiffPuter supports.

```{r}
library(RDiffPuter)
library(torch)
```

## Toy data

```{r}
set.seed(2)
n <- 300
truth <- data.frame(
  age      = pmax(18, rnorm(n, 40, 12)),
  income   = exp(rnorm(n, 10, 0.6)),
  edu      = factor(sample(c("hs", "ba", "ma", "phd"), n,
                           replace = TRUE, prob = c(0.4, 0.4, 0.15, 0.05))),
  region   = factor(sample(c("N", "S", "E", "W", "C"), n, replace = TRUE)),
  employed = sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(0.7, 0.3))
)

data <- truth
# 25% MCAR holes per column
for (col in names(data)) {
  miss <- sample.int(n, n * 0.25)
  data[miss, col] <- NA
}
sapply(data, function(x) sum(is.na(x)))
```

## Binary (DiffPuter-faithful) encoding

```{r}
device <- if (cuda_is_available()) "cuda" else
          if (torch::backends_mps_is_available()) "mps" else "cpu"

bin_run <- diffputer_em(
  data,
  encoding   = "binary",
  max_iter   = 3,
  epochs     = 300,
  patience   = 60,
  hidden_dim = 128,
  batch_size = 64,
  num_steps  = 20,
  num_trials = 8,
  num_resamplings = 5,
  device     = device,
  validation_data = truth,
  verbose    = TRUE
)
```

## One-hot encoding

The same call with `encoding = "onehot"` switches the categorical
representation. The encoded dimension grows because each level becomes its
own indicator column rather than one bit position.

```{r}
oh_run <- diffputer_em(
  data,
  encoding   = "onehot",
  max_iter   = 3,
  epochs     = 300,
  patience   = 60,
  hidden_dim = 128,
  batch_size = 64,
  num_steps  = 20,
  num_trials = 8,
  num_resamplings = 5,
  device     = device,
  validation_data = truth,
  verbose    = TRUE
)
```

## Compare

```{r}
# Categorical accuracy on the held-out cells
cat_acc <- function(imp) {
  sapply(c("edu", "region", "employed"), function(col) {
    miss <- is.na(data[[col]])
    if (!any(miss)) return(NA_real_)
    mean(as.character(imp[[col]][miss]) ==
           as.character(truth[[col]][miss]))
  })
}
rbind(binary = cat_acc(bin_run$imputed),
      onehot = cat_acc(oh_run$imputed))
```

`encoded_dim` for both:

```{r}
c(binary = bin_run$transformer$encoded_dim,
  onehot = oh_run$transformer$encoded_dim)
```
