---
title: "Multiple imputation with RDiffPuter"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple imputation with RDiffPuter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE,
                      purl = FALSE)
```

Single imputation gives you a point prediction for each missing cell —
useful for downstream prediction tasks, but it discards the model's
uncertainty about the missing values. For statistical inference you usually
want **multiple imputation**: produce `m` completed data sets, fit an
analysis model on each, and pool the results with Rubin's rules to recover
valid standard errors.

`RDiffPuter` exposes this through two functions:

- `multi_impute(model, data, transformer, m)` — given an already-trained
  model, draw `m` completed data sets.
- `diffputer_em_mi(data, m, ...)` — train and draw in one call.

```{r}
library(RDiffPuter)
library(torch)
device <- if (cuda_is_available()) "cuda" else
          if (torch::backends_mps_is_available()) "mps" else "cpu"
```

## A toy data set with missing entries

```{r}
set.seed(1)
n <- 300
truth <- data.frame(
  x1 = rnorm(n),
  x2 = rnorm(n),
  y  = NA_real_
)
truth$y <- 1.5 + 0.7 * truth$x1 - 0.3 * truth$x2 + rnorm(n, sd = 0.5)

data <- truth
data$x1[sample.int(n, 60)] <- NA
data$x2[sample.int(n, 60)] <- NA
```

## Train and draw with `diffputer_em_mi()`

```{r}
res <- diffputer_em_mi(
  data, m = 10,
  max_iter   = 3, epochs = 200, hidden_dim = 64,
  batch_size = 64, patience = 50,
  num_steps  = 20, num_resamplings = 5, num_trials = 4,
  mi_num_trials = 1,        # one Monte Carlo sample per completed dataset
  mi_num_steps  = 20,
  device     = device, verbose = TRUE
)
length(res$imputations)
print(res)
```

## Pool with `mice::pool()`

```{r}
fits <- lapply(res$imputations, function(d) lm(y ~ x1 + x2, data = d))
summary(mice::pool(fits))
```

Compare with the (over-confident) single-imputation result:

```{r}
em_single <- diffputer_em(
  data, max_iter = 3, epochs = 200, hidden_dim = 64,
  batch_size = 64, patience = 50,
  num_steps = 20, num_resamplings = 5, num_trials = 16,
  device = device, verbose = TRUE
)
summary(lm(y ~ x1 + x2, data = em_single$imputed))
```

The pooled standard errors will be larger — that's the point of MI. The
difference quantifies how much uncertainty the imputer was hiding.

## Reusing a trained model

If you already have a `RDiffPuter_em` object (e.g. from a previous run),
`multi_impute()` lets you draw additional data sets without retraining:

```{r}
mi_more <- multi_impute(em_single, data, m = 20, seed = 99,
                        num_steps = 20, num_resamplings = 5)
length(mi_more)
```

## Tuning the within-draw / between-draw trade-off

By default `num_trials = 1` per draw, which gives Rubin-style posterior
samples and faithful between-draw variance. Increasing `num_trials`
averages multiple sampling passes per completed dataset, which lowers
within-dataset noise but **also** reduces between-dataset variability and
can lead to optimistic pooled standard errors. Stick to `num_trials = 1`
unless you have a reason to deviate.
